Modelling state-switching functional data with hidden Markov models

### 2020-12-25 Report/Sections/simulation.tex
Nancy edits through 4.1

We generate 500 independent data sets using the CarHHMM-DFT as a generative model. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.79 & 0.21 \\ 0.81 & 0.19 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters are $\mu^{(1)} = 25.7s$, $\sigma^{(1)} = 9.6s$, $\mu^{(2)} = 104.6s$, and $\sigma^{(2)} = 64.7s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$. Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.034s, \enspace \phi_A^{*(\cdot,1)} = 0.98, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.079s, \enspace \phi_A^{*(\cdot,2)} = 0.87, \\
    %
    \mu_W^{*(\cdot,1)} = 23.3, \quad \sigma_W^{*(\cdot,1)} = 13.0, \\
    %
    \mu_W^{*(\cdot,2)} = 301.2, \quad \sigma_W^{*(\cdot,2)} = 330.1.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix. Figure 21 of the supplementary material shows one realization of $\Z$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

The two simulated dive types differ in that dives of type one are much shorter on average (26 seconds) than dives of type two (105 seconds). The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$ since both are much higher for subdive state 2 than for subdive state 1. These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data. All four candidate models are fit to each of the 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.

@@

We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}). If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives. Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows. We find both the average dive decoding accuracy and the average subdive decoding accuracy using the \textit{forward-backward} algorithm \citep{Zucchini:2016}.

>>>>

We generate 500 independent data sets using the CarHHMM-DFT as a generative model. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.79 & 0.21 \\ 0.81 & 0.19 \end{pmatrix}.$$
%
Dive duration is Gamma distributed and the coarse-scale emission parameters are 
%
\[\mu^{(1)} = 25.7s, \enspace \sigma^{(1)} = 9.6s, \enspace \mu^{(2)} = 104.6s, \enspace \sigma^{(2)} = 64.7s.\]
%
After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \right\}$ with probability transition matrices
%
\[\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}\]
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$. Recall from Section \ref{subsec:model_selection} that we must specify the mean, standard deviation, and auto-correlation parameters corresponding to $\left\{\Zone_{t,1},\ldots,\Zone_{t,\tilde T_t^*}\right\}$ as well as the mean and standard deviation parameters corresponding to $\left\{\Ztwo_{t,1},\ldots,\Ztwo_{t,\tilde T_t^*}\right\}$. We select the following parameters in line with the results from the case study:
%$\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.034s, \enspace \phi_A^{*(\cdot,1)} = 0.98, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.079s, \enspace \phi_A^{*(\cdot,2)} = 0.87, \\
    %
    \mu_W^{*(\cdot,1)} = 23.3, \quad \sigma_W^{*(\cdot,1)} = 13.0, \\
    %
    \mu_W^{*(\cdot,2)} = 301.2, \quad \sigma_W^{*(\cdot,2)} = 330.1.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix. Figure 21 of the supplementary material shows one realization of $\Z$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

The two simulated dive types differ in that dives of type 1 are much shorter on average (26 seconds) than dives of type 2 (105 seconds). The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$ since both are much higher for subdive state 2 than for subdive state 1. These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data. All four candidate models are fit to each of the 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.

@@

We use several metrics to compare the candidate models, including accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}). If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives. Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows. We find both the average dive decoding accuracy and the average subdive decoding accuracy using the \textit{forward-backward} algorithm \citep{Zucchini:2016}.
